---
title: "mvBIMBAM Installation Hack"
author: "Lacey W. Heinsberg"
output:
  github_document:
  html_preview: false
  toc: true
  pdf_document:
    toc: yes
    number_sections: true
    toc_depth: '5'
  html_document:
    df_print: paged
    toc: yes
    number_sections: true
    toc_depth: '5'
header-includes: 
  - \renewcommand{\and}{\\}
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
```

# Copyright information

Copyright 2024, University of Pittsburgh. All Rights Reserved. License: [GPL-2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)

# mvBIMBAM

## Mac

The motivation behind modifying the makefile was that during installation (of mvBIMBAM "first release"), test users were getting the following error:

```         
cd src && /Applications/Xcode.app/Contents/Developer/usr/bin/make
g++ -static-libgcc -DIMPUTATION   -O3 control.o fpmath.o indiv.o diploid.o haploid.o model.o param.o  fp.o -lm libgsl.a libgslcblas.a  -o bimbam
clang: error: unsupported option '-static-libgcc'
make[1]: *** [fp] Error 1
make: *** [all] Error 2
```

which suggests there is an issue with the build process. To get around this, we "hacked" a solution by modifying the Makefile (located within the `src` folder). See a summary below, or a "track changes" comparison in `makefile_compare_AppleIntel.docx` or `makefile_compare_AppleSilicon_M1.docx` (located here in the `InstallationAppendix` folder) for details on how the makefile was modified.

Note that the makefile changes are different for an Apple Silicon (e.g., 2021 M1) vs. an Apple Intel. If you are not sure which Mac you have, you can click on the apple symbol in the upper left corner of your machine. If you have an Apple silicon, the Chip will be listed as Apple M1 or M2. If you have the Apple Intel, you will see the processor listed as Intel Core i5, i7, or similar.

Note that this modified makefile and redistribution of mvBIMBAM software is redistributed as permitted under the creator's original GPL-3 license. 

### APPLE INTEL INSTRUCTIONS

**CHANGE A** Add GSL Compiler and Linker Flags

```         
CFLAGS += -I/usr/local/Cellar/gsl/2.7.1/include
LDFLAGS += -L/usr/local/Cellar/gsl/2.7.1/lib
```

These flags include the GSL headers and dynamically specify the location of the GSL libraries on a machine.

**CHANGE B** Update LIBS Variable. Specifically, modify:

```         
LIBS += -lm libgsl.a libgslcblas.a
```

to:

```         
LIBS += -lm -lgsl -lgslcblas
```

This directly links the GSL libraries for math, GSL core, and GSL CBLAS.

**CHANGE C** Update Compilation Command: Remove the `-static-libgcc` flag. Specifically, modify this line:

```         
fp: fp.o $(OBJS); $(CC) -static-libgcc $(CFLAGS) $(OBJS) fp.o $(LIBS) -o bimbam
```

to

```         
fp: fp.o $(OBJS); $(CC) $(CFLAGS) $(OBJS) fp.o $(LIBS) -o bimbam
```

The `-static-libgcc` flag is related to linking the GCC (GNU Compiler Collection) runtime libraries statically. However, this flag is not supported by the Clang compiler, which is commonly used on macOS (including Apple Silicon M1 Macs). By removing it, the build rule becomes compatible with Clang and more common macOS build setups.

A copy of the original Makefile, modified Makefile, and a Word document comparing the two can be found in the `Install_Problems` folder.

Note that this problem has been opened as an issue on the mvBIMBAM GitHub page <https://github.com/heejungshim/mvBIMBAM/issues/3>. Please check there for updates from the program architects.

### APPLE SILICON (M1)

**CHANGE A** Add GSL Compiler and Linker Flags

```         
CFLAGS += `gsl-config --cflags`
LIBS += `gsl-config --libs`
```

These flags include the GSL headers and dynamically specify the location of the GSL libraries on your machine.

**CHANGE B** Update LIBS Variable: Specifically, modify:

```         
LIBS += -lm libgsl.a libgslcblas.a
```

to:

```         
LIBS += -lm -lgsl -lgslcblas
```

This directly links the GSL libraries for math, GSL core, and GSL CBLAS.

**CHANGE C** Update Compilation Command: The compilation command for the "fp" target should be updated to remove the `-static-libgcc` flag and add a linker. Specifically, modify:

```         
fp: fp.o $(OBJS); $(CC) -static-libgcc $(CFLAGS) $(OBJS) fp.o $(LIBS) -o bimbam
```

to

```         
fp: fp.o $(OBJS); $(CC) $(CFLAGS) $(OBJS) fp.o $(LDFLAGS) $(LIBS) -o bimbam
```

The `-static-libgcc` flag is related to linking the GCC (GNU Compiler Collection) runtime libraries statically. However, this flag is not supported by the Clang compiler, which is commonly used on macOS (including Apple Silicon M1 Macs). By removing it, the build rule becomes compatible with Clang and more common macOS build setups.

Note also that Apple Silicon also requires the extra `$(LDFLAGS)` (where Intel does not) which is used to specify linker-related flags and options that may be specific to the architecture of M1 (not needed by Intel).

A copy of the original Makefile, modified Makefile, and a Word document comparing the two can be found in the `Install_Problems` folder.

Note that this problem has been opened as an issue on the mvBIMBAM GitHub page <https://github.com/heejungshim/mvBIMBAM/issues/3>. Please check there for updates from the program architects.

# Session information

```{r info}
sessioninfo::session_info()
```
